---
title: "Project_Week1_Statistics-GenomicDataScience"
author: "Haris Spyridis"
date: "2024-03-24"
output:
  html_document: default
  pdf_document: default
  fig_width: 15
---

```{r setup, eval=TRUE}
knitr::opts_chunk$set(cache=TRUE) 
```

```{r }
x = rnorm(10)
plot(x,pch=19,col="dodgerblue")
```


```{r }
y = rbinom(20,size=1,prob=0.5)
table(y)

?SummarizedExperiment
```




```{r}
library(Biobase)
library(GenomicRanges)
library(SummarizedExperiment)

data(sample.ExpressionSet, package = "Biobase")
se = makeSummarizedExperimentFromExpressionSet(sample.ExpressionSet)
colData(se)
rowRanges(se)
??summarizedExperiment
```


#Question 5:
'''Just considering the phenotype data what are some reasons that the Bottomly data set is likely a better experimental design than the Bodymap data? Imagine the question of interest in the Bottomly data is to compare strains and in the Bodymap data it is to compare tissues.'''

``` {r }
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bottomly_eset.RData")
load(file=con)
close(con)
bot = bottomly.eset
pdata_bot=pData(bot)
head(pdata_bot)

con = url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
pdata_bm=pData(bm)
head(pdata_bm)

```



#Question 6
"What are some reasons why this plot is not useful for comparing the number of technical replicates by tissue (you may need to install the plotrix package)."
``` {r}
library(plotrix)

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
pdata_bm=pData(bm)

pie3D(pdata_bm$num.tech.reps,labels=pdata_bm$tissue.type)
edata = exprs(bm)

#Answer: The "mixture" category is split across multiple wedges. 
```

#Question 7:
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)

row_sums = rowSums(edata)
edata = edata[order(-row_sums),]
index = 1:500
heatmap(edata[index,],Rowv=NA,Colv=NA)

```




#Question 8:
Make an MA-plot of the first sample versus the second sample using the log2 transform (hint: you may have to add 1 first) and the rlogrlog transform from the DESeq2 package. How are the two MA-plots different? Which kind of genes appear most different in each plot?
``` {r} 
library(DESeq2)

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
pdata = pData(bm)
edata = exprs(bm)
  
#MA plot with log2 transformation
edata_log = log2(edata+1)
mm_log = edata_log[,1] - edata_log[,2]
aa_log = edata_log[,1] + edata_log[,2]

plot(aa_log,mm_log)

#first perform rlog transformation for the whole dataset
edata_rlog = rlog(edata+1)

mmrlog = edata_rlog[,1] - edata_rlog[,2]
aarlog= edata_rlog[,1] + edata_rlog[,2]
plot(aarlog, mmrlog)

```




Question 9 : Load the Montgomery and Pickrell eSet
```{r}

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)



```
Cluster the data in three ways:

  1)With no changes to the data

  2)After filtering all genes with rowMeans  less than 100

  3)After taking the log2log2 transform of the data without filtering

#clust with no changes
``` {r fig.width=10, fig.height=6}
library(rafalib)
dist1 = dist(t(edata))
hclust1 = hclust(dist1)
col<-c(rep(2,60), rep(3, 69)) # 2 and 3 indicate red nd green
myplclust(hclust = hclust1, labels = hclust1$labels, lab.col = col)

```

#after filtering all genes with rowmeans<100
```{r fig.width=10, fig.height=6}
filter_means = edata[rowMeans(edata)<100, ]
dist2 = dist(t(filter_means))
hclust2 = hclust(dist2)
myplclust(hclust = hclust2, labels = hclust2$labels, lab.col = col)


```

#with log2 transformation without filtering
```{r fig.width=10, fig.height=6}
filter_log = log2(edata+1)
dist3 = dist(t(filter_log))
hclust3 = hclust(dist3)
myplclust(hclust = hclust3, labels = hclust3$labels, lab.col = col)


```






#Question 10
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata= as.data.frame(exprs(mp))
fdata = fData(mp)
dist4 = dist(t(filter_log))


filter_log = log2(edata+1)

kmeans_1 = kmeans(t(filter_log), centers = 2)
#This is particularly useful when you want to ensure that your results are reproducible,
set.seed(12345)
matplot(t(kmeans1$centers), col=1:2, type="l", lwd = 2)
kmeans_labels = kmeans_1$cluster
table(kmeans_labels)


#cut tree but first we need hierarchicalc lust
dist4 = dist(t(filter_log))
hclust4 = hclust(dist4)
hc = cutree(hclust4, k=2)
table(cluster_labels, kmeans_labels)
plot(hc)

dend = color_labels(hclust4, 2,1:2)
plot(dend)

?cutree



```
